On the power of conditional independence testing \\ under model-X

### 2021-07-23 crtpower.tex
aadi pass

\addbibresource{~/.references/crtpower}

@@

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (assuming $\text{dim}(\prx) = 1$) in a cross-validated lasso~\cite{T96} of $\sry$ on $\srx$ and $\srz$ \cite{CetL16} . To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, we resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which we extract the upper quantile
\begin{equation}
	C_{\alpha}(\sry,\srz) \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz],
	\label{upper-quantile}
\end{equation}
where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Then, the CRT rejects if the original test statistic exceeds this quantile:

\begin{equation}
	\label{eq:randomized-CRT}
	\phi^{\CRT}_T(\srx,\sry,\srz) \equiv 
	\begin{cases}
		1, \quad &\text{if }  T(\srx, \sry, \srz) > C_\alpha(\sry,\srz); \\
		\gamma, \quad &\text{if }  T(\srx, \sry, \srz) = C_\alpha(\sry,\srz); \\
		0, \quad &\text{if }T(\srx, \sry, \srz) < C_\alpha(\sry,\srz).
	\end{cases}
\end{equation}
In order to accommodate discreteness, the CRT makes a randomized decision $\gamma$ when $T(\srx, \sry, \srz) = C_\alpha(\sry,\srz)$ so that the size of the test is exactly $\alpha$.
%rejects with probability \[\gamma \equiv \frac{\alpha - \mathbb {P}[T(\srxk, \sry, \srz) > C_\alpha(\sry,\srz)|\sry,\srz]}{\mathbb {P}[T(\srxk, \sry, \srz) = C_\alpha(\sry,\srz)|\sry,\srz]}\] in cases when the test statistic is exactly equal to the rejection threshold. Note that the above probabilities in the definition of $\gamma$ are over the resampling distribution $\srxk\ |\ \sry, \srz$. 
In practice, the threshold $C_\alpha(\sry,\srz)$ is approximated by computing $T(\srxk^b, \sry, \srz)$ for a large number $B$ of Monte Carlo resamples $\srxk^b \sim \srx|\srz$. % For $b = 1,\dots, B$, a resample $\srxk^b$ is drawn and the statistic $T(\srxk^b, \sry, \srz)$ is recomputed. Then, the CRT threshold is obtained through the empirical quantile
%\begin{equation}
%	\widehat C_\alpha(\sry, \srz) \equiv Q_{1-\alpha}\left\{T(\srx, \sry, \srz), \{T(\srxk^b, \sry, \srz)\}_{b = 1, \dots, B}\right\}.
%	\label{empirical-quantile}
%\end{equation}
%If desired, a size $\alpha$ test can be obtained by computing $\gamma$ in the spirit of definition~\eqref{empirical-quantile}, but usually the CRT is implemented as a non-randomized test rejecting when $T(\srx, \sry, \srz) > \widehat C_\alpha(\sry, \srz)$. 
%The randomization $\gamma$ is not applied in practice, the test rejecting when $T(\srx, \sry, \srz) > \widehat C_\alpha(\sry, \srz)$. The finite-sample validity of the CRT for any $B \leq \infty$ and any $n,d,r,p$ follows from the fact that the resampled triples $(\srxk, \sry, \srz)$ are exchangeable with the original $(\srx, \sry, \srz)$ under the null, much like the argument used for a permutation test. 
For the sake of clarity, in this paper we consider only the ``infinite-$B$" version of the CRT as defined by equations~\eqref{upper-quantile} and \eqref{eq:randomized-CRT}. In the causal inference setting, the CRT can be viewed as a variant of Fisher's exact test for randomized experiments that incorporates strata of covariates \cite{Zheng2008,Hennessy2016}, basing inference on rerandomizing the treatment to the units.

The conditional independence testing problem under MX has benefited from a variety of methodological innovations, but there are still many open theoretical questions, including the following three:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] To what extent can the MX assumption be weakened?
	\item[Q3.] What is the precise connection between the performance of the machine learning (ML) algorithm and the power of the resulting MX method?
\end{enumerate}
In this paper, we shed light on these questions. We summarize our findings next.

@@

Let us first formalize the definition of a level $\alpha$ test of the MX CI problem. The null hypothesis is defined as the set of joint distributions compatible with conditional independence and with the assumed model for $\prx|\prz$:
\begin{equation}
	\label{eq:null-under-modelX}
	\begin{split}
		\mathscr L^{\text{MX}}_0(f^*) &\equiv \mathscr L_0 \cap \mathscr L^{\text{MX}}(f^*) \\
		&\equiv \{\mathcal L: \prx \independent \pry \mid \prz\} \cap \{\mathcal L: \mathcal L(\prx|\prz) = f^*_{\prx|\prz}\}\\
		&= \{\mathcal L: \mathcal L(\prx,\pry,\prz) = f_{\prz} \cdot f^*_{\prx|\prz} \cdot f_{\pry|\prz} \text{ for some } f_{\prz}, f_{\pry|\prz} \}.
	\end{split}
\end{equation}
A test $\phi: (\mathbb R^{d} \times \mathbb R^r \times \mathbb R^p)^n \rightarrow [0,1]$ of the MX CI problem is level $\alpha$ if
\begin{equation}
	\sup_{\mathcal L \in \mathscr L^{\text{MX}}_0(f^*)} \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)] \leq \alpha.
	\label{marginal-validity}
\end{equation}

Since the CRT calibrates the test statistic $T$ conditionally on the observed $\sry, \srz$ (recall definition~\eqref{upper-quantile}), it is easy to verify that this test not only has level $\alpha$ in the sense of equation~\eqref{marginal-validity} but also has level $\alpha$ \textit{conditionally} on $\sry$ and $\srz$:
\begin{equation}
	\sup_{y,z} \sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi^{\textnormal{CRT}}_T(\srx, \sry, \srz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
\label{eq:conditional-validity}
\end{equation}
It turns out that any level $\alpha$ test $\phi$ has this same property:\footnote{We thank Michael Celentano for pointing out this fact to us.}
\begin{proposition} \label{prop:marginal-implies-conditional}
	Any level $\alpha$ test $\phi$ of conditional independence under the MX assumption (i.e. any test satisfying property~\eqref{marginal-validity}) must also have conditional level $\alpha$:
	\begin{equation}
		\sup_{y,z} \sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
	\end{equation}
\end{proposition}

@@

The preceding observations will not be surprising to anyone familiar with MX methodology, and in fact the existence of a single null distribution from which to resample $\srxk$ is central to the very definition of the CRT. Nevertheless, we find it useful to state explicitly what has thus far been largely left implicit. This conditional perspective allows us to derive the optimal test against a point alternative by applying the  Neyman-Pearson lemma in the conditional problem.

@@

Theorem~\ref{prop:crt-optimality} and Corollary~\ref{cor:crt-optinality} state that the most powerful test against a point alternative is the CRT based on the test statistic defined as the measuring how well the data $(\srx, \sry, \srz)$ fit the distribution $\bar{\mathcal L}(\pry|\prx,\prz)$. For example, if
\begin{equation}
\bar f(\pry|\prx,\prz) = N(\prx^T\beta + \prz^T \gamma, \sigma^2) \text{ for coefficients } \beta \in \mathbb R^d \text{ and } \gamma \in \mathbb R^p,
\label{linear-model}
\end{equation}
then the optimal test rejects for small values of $\|\sry - \srx \beta - \srz \gamma \|^2$. In Section~\ref{sec:knockoffs}, we establish an analogous optimality statement for MX knockoffs as well. Since the optimal test depends on the alternative distribution $\bar{\mathcal L}(\pry|\prx,\prz)$, CRT and MX knockoffs implementations usually employ a machine learning step to search through the composite alternative (not unlike a likelihood ratio test) for a good approximation $\widehat f_{\pry|\prx,\prz}$. These approximate models are then summarized in various ways to define a test statistic $T$. There is no consensus yet on the best test statistic to use, with some authors \cite{CetL16, SetC17, SetS19} using combinations of fitted coefficients $\widehat \beta$ and others  \cite{Tansey2018, Bates2020} using loss-based test statistics. The above optimality results align more closely with the latter strategy. Loss-based test statistics also have the advantage of avoiding ad hoc combination rules for $\widehat \beta \in \mathbb R^d$ in cases where $d > 1$. It remains to be seen whether loss-based or coefficient-based test statistics yield greater power in practice.

@@

\section{Weakening the MX assumption} \label{sec:weakening}

@@

For each $n = 1, 2, \dots$, we have a joint law $\mathcal L_n$ over $(\prx, \pry, \prz) \in \mathbb R^{d + r + p}$, where $d = \text{dim}(\prx)$ remains fixed, $r = \text{dim}(\pry) = 1$, and $p = \text{dim}(\prz)$ can vary arbitrarily with $n$. Under the MX(2) assumption, the law $\mathcal L_n$ has known mean and variance functions
\begin{equation}
	\mu_n(\prz) \equiv \mathbb E_{\mathcal L_n}[\prx|\prz]  \text{ and } \Sigma_n(\prz) \equiv \text{Var}_{\mathcal L_n}[\prx|\prz].
	\label{conditional-mean-variance}
\end{equation}
For each $n$, we receive $n$ i.i.d. samples $(\srx, \sry, \srz) = \{(\srx_i, \sry_i, \srz_i)\}_{i = 1}^n$ from $\mathcal L_n$. Note that we leave implicit the dependence on $n$ of $(\prx,\pry,\prz)$ and $(X,Y,Z)$ to lighten the notation.

@@

Suppose we have trained an estimate $\widehat g_n$ of $\mathbb E_{\mathcal L_n}[\pry|\prz]$ on an independent dataset (whose size can vary arbitrarily with $n$ and is not included in the sample size $n$ used for testing). In the next section, $g_n$ will denote be the nonparametric portion of a semiparametric model~\eqref{eq:linearity}. \textit{These training sets across $n$ and resulting estimates $\widehat g_n$ remain fixed throughout.} Importantly, the sample used for training $\widehat g_n$ need not come from the same distribution as $\mathcal L_n$ and can therefore be much larger. For example, in the context of the digital twin test \cite{Bates2020} it was pointed out that large observational datasets can be used to train $\widehat f_{\pry|\prx,\prz}$, while applying a variant of the CRT to a smaller experimental dataset to obtain a causal guarantee.

@@

For uniform type-I error control, we must restrict the set of null distributions~\eqref{mx2-null} to those satisfying the following moment conditions for fixed $c_1, c_2 > 0$:
\begin{equation}
	\mathscr L_n(c_1, c_2) \equiv \{\mathcal L_n: \|S_{n}^{-1}\| \leq c_1, \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^{4} \mathbb E_{\mathcal L_n}[\|\prx - \mu_n(\prz)\|^{4}|\prz]\right] \leq c_2 \}.
\end{equation}

@@

While the MX(2) $F$-test is quite different from usual MX methods on its surface, the next theorem states that it is asymptotically equivalent to the CRT based on the same test statistic.
\begin{theorem} \label{thm:equivalence}
Under Setting~\ref{setting:asymptotic}, suppose
	\begin{equation}
		\mathcal L_n \in \mathscr L^{\textnormal{MX(2)}}(\mu_n(\cdot), \Sigma_n(\cdot)) \cap \mathscr L_n(c_1,c_2)
	\end{equation}
	for some $c_1, c_2 > 0$, and define $T_n(\srx, \sry, \srz)$ via equations~\eqref{u-hat} and~\eqref{t-n}. Let $\phi_n^{\textnormal{CRT}}$ denote the CRT based on $T_n$, with threshold $C_n(\sry, \srz)$ defined as in equation~\eqref{upper-quantile}. The CRT threshold converges in probability to the MX(2) threshold:
	\begin{equation}
		C_n(Y,Z) \overset{\mathcal L_n}\rightarrow_p c_{d,1-\alpha}.
		\label{eq:threshold-convergence}
	\end{equation}
Furthermore, if $T_n(\srx, \sry, \srz)$ does not accumulate near $c_{d,1-\alpha}$, i.e.
		\begin{equation}
		\lim_{\delta \rightarrow 0}\limsup_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[|T_n(\srx, \sry, \srz)-c_{d,1-\alpha}| \leq \delta] = 0,
		\label{eq:non-accumulation}
		\end{equation}
	 	then the CRT is asymptotically equivalent to the MX(2) F-test:
		\begin{equation}
			\lim_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[\phi^{\textnormal{MX(2)}}_n(\srx, \sry, \srz) \neq \phi^{\textnormal{CRT}}_n(\srx, \sry, \srz)] = 0.
			\label{eq:asymptotic-equivalence}
		\end{equation}	
\end{theorem}

Informally, this theorem suggests that the null distribution of $U_n(\srx, \sry, \srz)$ converges to $N(0, I_d)$, even after conditioning on $\sry, \srz$. In the language of the CRT, this means that the resampling distribution of the test statistic $T_n(\srx, \sry, \srz)$ approaches $\chi^2_d$. This is not too surprising in retrospect, since $\text{Var}_{\mathcal L_n}[\sqrt n \widehat \rho_n | \sry, \srz] = \widehat S_n^2$.  Note that this conditional normalization property holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior. We can also view Theorem~\ref{thm:equivalence} as a robustness result. Indeed, this theorem implies that the large-sample behavior of the CRT based on $T_n$ depends only on the first two moments of $\prx|\prz$. Therefore, the CRT carried out using a misspecified distribution $\prx|\prz$ that is correct up to at least the first two moments will be asymptotically equivalent to the CRT based on the correct distribution and therefore, will control Type-I error asymptotically as well.

@@

Theorem~\ref{thm:equivalence} is a statement about the asymptotic equivalence between the resampling-based CRT and the asymptotic MX(2) $F$-test. The CRT is in the spirit of the finite-population approach to causal inference (Fisher), whereas the MX(2) $F$-test is in the spirit of the asymptotic super-population approach (Neyman). We find that research in these two strands of work on causal inference have proceeded largely separately from each other, and therefore connections between the two have received relatively little attention. However, there has been a recent line of work \cite{Ding2017,Wu2020a,Zhao2021} focusing on the asymptotic behavior of the Fisher randomization test in the context of completely randomized experiments. A similar result to Theorem~\ref{thm:equivalence} is that the Fisher randomization test (analogous to the CRT) is asymptotically equivalent to the Rao score test (analogous to the MX(2) $F$-test) in a completely randomized experiment \cite[Theorem A.1]{Ding2017}. Theorem~\ref{thm:equivalence} can be viewed as an extension of this result to accommodate for non-binary treatments as well as high-dimensional covariates affecting both treatment and response.

@@

Note that, if $(\prx,\prz)$ is jointly Gaussian, then $U_n(\srx, \sry, \srz)$ is exactly distributed as $N(0,I_d)$ both unconditionally and conditionally in finite samples; see also Section 2.5 of \cite{Liu2020}. To test the above convergence statements in a nontrivial setting, we instead consider a discrete distribution for $(\prx,\prz)$. In particular, we sample $(\prx, \prz)$ from a Markov chain, as described next.

>>>>

%\addbibresource{~/.references/crtpower}
\addbibresource{crtpower}

@@

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (when $\text{dim}(\prx) = 1$) in a cross-validated lasso~\cite{T96} of $\sry$ on $\srx$ and $\srz$ \cite{CetL16}. To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which the upper quantile
\begin{equation}
	C_{\alpha}(\sry,\srz) \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz]
	\label{upper-quantile}
\end{equation}
may be extracted,
where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Finally, the CRT rejects if the original test statistic exceeds this quantile:

\begin{equation}
	\label{eq:randomized-CRT}
	\phi^{\CRT}_T(\srx,\sry,\srz) \equiv 
	\begin{cases}
		1, \quad &\text{if }  T(\srx, \sry, \srz) > C_\alpha(\sry,\srz); \\
		\gamma, \quad &\text{if }  T(\srx, \sry, \srz) = C_\alpha(\sry,\srz); \\
		0, \quad &\text{if }T(\srx, \sry, \srz) < C_\alpha(\sry,\srz).
	\end{cases}
\end{equation}
In order to accommodate discreteness, the CRT makes a randomized decision $\gamma$ when $T(\srx, \sry, \srz) = C_\alpha(\sry,\srz)$ so that the size of the test is exactly $\alpha$.
%rejects with probability \[\gamma \equiv \frac{\alpha - \mathbb {P}[T(\srxk, \sry, \srz) > C_\alpha(\sry,\srz)|\sry,\srz]}{\mathbb {P}[T(\srxk, \sry, \srz) = C_\alpha(\sry,\srz)|\sry,\srz]}\] in cases when the test statistic is exactly equal to the rejection threshold. Note that the above probabilities in the definition of $\gamma$ are over the resampling distribution $\srxk\ |\ \sry, \srz$. 
In practice, the threshold $C_\alpha(\sry,\srz)$ is approximated by computing $T(\srxk^b, \sry, \srz)$ for a large number $B$ of Monte Carlo resamples $\srxk^b \sim \srx|\srz$. % For $b = 1,\dots, B$, a resample $\srxk^b$ is drawn and the statistic $T(\srxk^b, \sry, \srz)$ is recomputed. Then, the CRT threshold is obtained through the empirical quantile
%\begin{equation}
%	\widehat C_\alpha(\sry, \srz) \equiv Q_{1-\alpha}\left\{T(\srx, \sry, \srz), \{T(\srxk^b, \sry, \srz)\}_{b = 1, \dots, B}\right\}.
%	\label{empirical-quantile}
%\end{equation}
%If desired, a size $\alpha$ test can be obtained by computing $\gamma$ in the spirit of definition~\eqref{empirical-quantile}, but usually the CRT is implemented as a non-randomized test rejecting when $T(\srx, \sry, \srz) > \widehat C_\alpha(\sry, \srz)$. 
%The randomization $\gamma$ is not applied in practice, the test rejecting when $T(\srx, \sry, \srz) > \widehat C_\alpha(\sry, \srz)$. The finite-sample validity of the CRT for any $B \leq \infty$ and any $n,d,r,p$ follows from the fact that the resampled triples $(\srxk, \sry, \srz)$ are exchangeable with the original $(\srx, \sry, \srz)$ under the null, much like the argument used for a permutation test. 
For the sake of clarity, this paper considers only the ``infinite-$B$" version of the CRT as defined by~\eqref{upper-quantile} and \eqref{eq:randomized-CRT}. In the causal inference setting, the CRT can be viewed as a variant of Fisher's exact test for randomized experiments that incorporates strata of covariates \cite{Zheng2008,Hennessy2016}, basing inference on rerandomizing the treatment to the units.

The CI testing problem under MX has benefited from several methodological innovations, but there are still many open theoretical questions, including the following three:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] To what extent can the MX assumption be weakened?
	\item[Q3.] What is the precise connection between the performance of the machine learning (ML) algorithm and the power of the resulting MX method?
\end{enumerate}
In this paper, we shed light on these questions. We summarize our findings next.

@@

Let us first formalize the definition of a level $\alpha$ test of the MX CI problem. The null hypothesis is defined as the set of joint distributions compatible with conditional independence and with the assumed model for $\prx|\prz$:
\begin{equation}
	\label{eq:null-under-modelX}
	\begin{split}
		\mathscr L^{\text{MX}}_0(f^*) &\equiv \mathscr L_0 \cap \mathscr L^{\text{MX}}(f^*) \\
		&\equiv \{\mathcal L: \prx \independent \pry \mid \prz\} \cap \{\mathcal L: \mathcal L(\prx|\prz) = f^*_{\prx|\prz}\}\\
		&= \{\mathcal L: \mathcal L(\prx,\pry,\prz) = f_{\prz} \cdot f^*_{\prx|\prz} \cdot f_{\pry|\prz} \text{ for some } f_{\prz}, f_{\pry|\prz} \}.
	\end{split}
\end{equation}
A test $\phi: (\mathbb R^{d} \times \mathbb R^r \times \mathbb R^p)^n \rightarrow [0,1]$ of the MX CI problem is said to be level $\alpha$ if
\begin{equation}
	\sup_{\mathcal L \in \mathscr L^{\text{MX}}_0(f^*)} \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)] \leq \alpha.
	\label{marginal-validity}
\end{equation}

Since the CRT calibrates the test statistic $T$ conditionally on the observed $\sry, \srz$ (recall definition~\eqref{upper-quantile}), it is easy to verify that this test not only has level $\alpha$ in the sense of equation~\eqref{marginal-validity} but also has level $\alpha$ \textit{conditionally} on $\sry$ and $\srz$:
\begin{equation}
	\sup_{y,z} \sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi^{\textnormal{CRT}}_T(\srx, \sry, \srz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
\label{eq:conditional-validity}
\end{equation}
It turns out that, not just the CRT, but \emph{any} level $\alpha$ test $\phi$ has this same property:\footnote{We thank Michael Celentano for pointing out this fact to us.}
\begin{proposition} \label{prop:marginal-implies-conditional}
	Any level $\alpha$ test $\phi$ of conditional independence under the MX assumption (i.e. any test satisfying property~\eqref{marginal-validity}) must also be conditionally level $\alpha$:
	\begin{equation}
		\sup_{y,z} \sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
	\end{equation}
\end{proposition}

@@

The preceding observations may not come as a surprise to those familiar with MX methodology, and in fact the existence of a single null distribution from which to resample $\srxk$ is central to the very definition of the CRT. Nevertheless, we find it useful to state explicitly what has thus far been largely left implicit. This conditional perspective allows us to derive the optimal test against a point alternative by applying the  Neyman-Pearson lemma in the conditional problem.

@@

Theorem~\ref{prop:crt-optimality} and Corollary~\ref{cor:crt-optinality} state that the most powerful test against a point alternative is the CRT based on the test statistic defined as the measuring how well the data $(\srx, \sry, \srz)$ fit the distribution $\bar{\mathcal L}(\pry|\prx,\prz)$. For example, if
\begin{equation}
\bar f(\pry|\prx,\prz) = N(\prx^T\beta + \prz^T \gamma, \sigma^2) \text{ for coefficients } \beta \in \mathbb R^d \text{ and } \gamma \in \mathbb R^p,
\label{linear-model}
\end{equation}
then the optimal test rejects for small values of $\|\sry - \srx \beta - \srz \gamma \|^2$. In Section~\ref{sec:knockoffs}, we establish an analogous optimality statement for MX knockoffs as well. Since the optimal test depends on the alternative distribution $\bar{\mathcal L}(\pry|\prx,\prz)$, CRT and MX knockoffs implementations usually employ a machine learning step to search through the composite alternative (not unlike a likelihood ratio test) for a good approximation $\widehat f_{\pry|\prx,\prz}$. These approximate models are then summarized in various ways to define a test statistic $T$. There is no consensus yet on the best test statistic to use, with some authors \cite{CetL16, SetC17, SetS19} using combinations of fitted coefficients $\widehat \beta$ and others  \cite{Tansey2018, Bates2020} using loss-based test statistics. The above optimality results align more closely with the latter strategy. Loss-based test statistics also have the advantage of avoiding ad hoc combination rules for $\widehat \beta \in \mathbb R^d$ when $d > 1$. 
It remains to be seen whether loss-based or coefficient-based test statistics yield greater power in practice, but a thorough empirical comparison is beyond the scope of this work. For now, it suffices to note that, despite its simplicity, this is the first such power optimality result in the CRT literature.

@@

\section{Weakening the MX assumption to MX(2)} \label{sec:weakening}

@@

Following a triangular array formalization, for each $n = 1, 2, \dots$, we have a joint law $\mathcal L_n$ over $(\prx, \pry, \prz) \in \mathbb R^{d + r + p}$, where $d = \text{dim}(\prx)$ remains fixed, $r = \text{dim}(\pry) = 1$, and $p = \text{dim}(\prz)$ can vary arbitrarily with $n$. Under the MX(2) assumption, the law $\mathcal L_n$ has known mean and variance functions
\begin{equation}
	\mu_n(\prz) \equiv \mathbb E_{\mathcal L_n}[\prx|\prz]  \text{ and } \Sigma_n(\prz) \equiv \text{Var}_{\mathcal L_n}[\prx|\prz].
	\label{conditional-mean-variance}
\end{equation}
For each $n$, we receive $n$ i.i.d. samples $(\srx, \sry, \srz) = \{(\srx_i, \sry_i, \srz_i)\}_{i = 1}^n$ from $\mathcal L_n$. Note that we leave implicit the dependence on $n$ of $(\prx,\pry,\prz)$ and $(X,Y,Z)$ to lighten the notation.

@@

Suppose we have trained an estimate $\widehat g_n$ of $\mathbb E_{\mathcal L_n}[\pry|\prz]$ on an independent dataset (whose size can vary arbitrarily with $n$ and is not included in the sample size $n$ used for testing). In the next section, $g_n$ will denote be the nonparametric portion of a semiparametric model~\eqref{eq:linearity}. These training sets across $n$ and resulting estimates $\widehat g_n$ remain fixed throughout. Importantly, the sample used for training $\widehat g_n$ \emph{need not come from the same distribution} as $\mathcal L_n$ and can therefore be much larger. For example, in the context of the digital twin test \cite{Bates2020} it was pointed out that large observational datasets can be used to train $\widehat f_{\pry|\prx,\prz}$, while applying a variant of the CRT to a smaller (randomized) experimental dataset to obtain a causal guarantee. Hence, our setting could be viewed from the lens of sample splitting, or in the spirit of drawing inferences when having a (typically large) observational dataset and a (likely small) experimental dataset. Both viewpoints will be useful in later sections.

@@

For uniform type-I error control, we must restrict the set of null distributions~\eqref{mx2-null} to those satisfying the following moment conditions for fixed $c_1, c_2 > 0$:
\begin{equation}
	\mathscr L_n(c_1, c_2) \equiv \{\mathcal L_n: \|S_{n}^{-1}\| \leq c_1, \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^{4} \mathbb E_{\mathcal L_n}[\|\prx - \mu_n(\prz)\|^{4}|\prz]\right] \leq c_2 \}.
\end{equation} 
For pointwise error control (less desirable), these moment conditions can be relaxed, but they are unavoidable for uniform type-1 error control.

@@

While the MX(2) $F$-test is quite different from usual MX methods on its surface, the next theorem states that it is asymptotically equivalent to the CRT (which requires the full MX assumption to run in practice) based on the same test statistic.
\begin{theorem} \label{thm:equivalence}
Under Setting~\ref{setting:asymptotic}, suppose
	\begin{equation}
		\mathcal L_n \in \mathscr L^{\textnormal{MX(2)}}(\mu_n(\cdot), \Sigma_n(\cdot)) \cap \mathscr L_n(c_1,c_2)
	\end{equation}
	for some $c_1, c_2 > 0$, and define $T_n(\srx, \sry, \srz)$ via equations~\eqref{u-hat} and~\eqref{t-n}. Let $\phi_n^{\textnormal{CRT}}$ denote the CRT based on $T_n$, with threshold $C_n(\sry, \srz)$ defined as in equation~\eqref{upper-quantile}. The CRT threshold converges in probability to the MX(2) threshold:
	\begin{equation}
		C_n(Y,Z) \overset{\mathcal L_n}\rightarrow_p c_{d,1-\alpha}.
		\label{eq:threshold-convergence}
	\end{equation}
Furthermore, if $T_n(\srx, \sry, \srz)$ does not accumulate near $c_{d,1-\alpha}$, i.e.
		\begin{equation}
		\lim_{\delta \rightarrow 0}\limsup_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[|T_n(\srx, \sry, \srz)-c_{d,1-\alpha}| \leq \delta] = 0,
		\label{eq:non-accumulation}
		\end{equation}
	 	then the CRT is asymptotically equivalent to the MX(2) F-test:
		\begin{equation}
			\lim_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[\phi^{\textnormal{MX(2)}}_n(\srx, \sry, \srz) \neq \phi^{\textnormal{CRT}}_n(\srx, \sry, \srz)] = 0.
			\label{eq:asymptotic-equivalence}
		\end{equation}	
\end{theorem}

Informally, this theorem suggests that the null distribution of $U_n(\srx, \sry, \srz)$ converges to $N(0, I_d)$, even after conditioning on $\sry, \srz$. In the language of the CRT, this means that the resampling distribution of the test statistic $T_n(\srx, \sry, \srz)$ approaches $\chi^2_d$. This is not too surprising in retrospect, since $\text{Var}_{\mathcal L_n}[\sqrt n \widehat \rho_n | \sry, \srz] = \widehat S_n^2$.  Note that this conditional normalization property holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior.

We can alternately view Theorem~\ref{thm:equivalence} as a robustness result for the CRT. Indeed, this theorem implies that the large-sample behavior of the CRT based on $T_n$ depends only on the first two moments of $\prx|\prz$. Therefore, the CRT carried out using a misspecified distribution $\prx|\prz$ that is correct up to at least the first two moments will be asymptotically equivalent to the CRT based on the correct distribution and therefore, will control Type-I error asymptotically as well.

@@

Theorem~\ref{thm:equivalence} is a statement about the asymptotic equivalence between the resampling-based CRT and the asymptotic MX(2) $F$-test. The MX CRT is in the spirit of the finite-population approach to causal inference (Fisher), whereas the MX(2) $F$-test is in the spirit of the asymptotic super-population approach (Neyman). We find that research in these two strands of work on causal inference have proceeded largely separately from each other, and therefore connections between the two have received relatively little attention. However, there has been a recent line of work \cite{Ding2017,Wu2020a,Zhao2021} focusing on the asymptotic behavior of the Fisher randomization test in the context of completely randomized experiments. A similar result to Theorem~\ref{thm:equivalence} is that the Fisher randomization test (analogous to the CRT) is asymptotically equivalent to the Rao score test (analogous to the MX(2) $F$-test) in a completely randomized experiment \cite[Theorem A.1]{Ding2017}. Theorem~\ref{thm:equivalence} can be viewed as an extension of this result to accommodate for non-binary treatments as well as high-dimensional covariates affecting both treatment and response.

@@

Note that, if $(\prx,\prz)$ is jointly Gaussian, then $U_n(\srx, \sry, \srz)$ is exactly distributed as $N(0,I_d)$ both unconditionally and conditionally in finite samples; see also Section 2.5 of \cite{Liu2020}. To test the above convergence statements in a nontrivial setting, we instead consider a discrete distribution for $(\prx,\prz)$. In particular, we sample $(\prx, \prz)$ from a Markov chain, as described next. (Such a Markovian setup has often been employed in MX analyses of population GWAS studies {\color{red}CITE}, motivated by genetic recombination models from theoretical biology.)

Let's assume for simplicity that $\text{dim}(\prx) = 1$. Define $(\prx,\prz) \in \{0,1\}^{1+p}$ to have the distribution of a Markov chain with 
\begin{equation*}
	\text{initial state } \prx \sim \text{Ber}(\pi_{\text{init}}) \text{ and transition matrix } \begin{pmatrix}1-\pi_\text{flip} & \pi_{\text{flip}} \\  \pi_{\text{flip}} &  1-\pi_{\text{flip}}\end{pmatrix}.
\end{equation*}
More explicitly, we have
\begin{equation*}
	\mathbb P[\prx = x, \prz = z] = \pi_{\text{init}}^{x}(1-\pi_{\text{init}})^{1-x} \pi_{\text{flip}}^{\mathbbm 1(z_1 \neq x)} (1-\pi_{\text{flip}})^{\mathbbm 1(z_1 = x)} \prod_{j = 2}^p \pi_{\text{flip}}^{\mathbbm 1(z_j \neq z_{j-1})} (1-\pi_{\text{flip}})^{\mathbbm 1(z_j = z_{j-1})}.
\end{equation*}
The parameters $(\pi_{\text{init}}, \pi_{\text{flip}})$ describe the distribution of $\prx|\prz$ and are assumed known. Furthermore, let the response $\pry$ be distributed as a random effects model in $\prz$:
\begin{equation*}
	\pry = \prz^T \bm \gamma + \peps, \quad \bm \gamma \sim N(0, \sigma^2_{\gamma}I_p),\ \peps \sim N(0, \sigma^2_\eps I_n).
\end{equation*}
Thus, all simulations are conducted under the null hypothesis $H_0: \prx \independent \pry \mid \prz$. The signal-to-noise ratio in this relationship is defined via
\begin{equation*}
	\text{SNR} = \frac{\mathbb E[\|\prz\|^2]\sigma^2_\gamma}{\sigma^2_\eps}.
\end{equation*}
The function $\widehat g_n$ is defined by running a 10-fold cross-validated ridge regression of $\sry$ on $\srz$ on $n_{\text{train}}$ training samples, and then the statistic $U_n(\srx, \sry, \srz)$ is evaluated based on $n_{\text{test}}$ independent test samples.
